Iterative Monomorphisation

### 2024-03-08 iterative_monomorphisation.tex
modifs with Tanguy

@@

@@

   \KwData{finite set \(\Phi\) of polymorphic formulae}
   \KwResult{finite set of monomorphic formulae}

@@

   \((M_{\text{old}}, P_{\text{old}}) \leftarrow (\emptyset, \emptyset)\)\;
   \((M_{\text{new}}, P_{\text{new}}) \leftarrow \Init(\Problem)\)\;

@@

   \For{\(i = 1\) \To \textcolor{ourblueviolet}{\Loop}}{
      \((M_{\text{next}}, P_{\text{next}})\leftarrow(\emptyset,\emptyset)\)\;
      \ForEach{\(\varphi \in \Problem\)} {
         \((M_{\Delta}, P_{\Delta}(\varphi), S_{\Delta})\leftarrow \FStep(M_{\text{old}}, P_{\text{old}}(\varphi), M_{\text{new}}, P_{\text{new}}(\varphi))\)\;
         \(\AllSubst\leftarrow\AllSubst\cup S_{\Delta}\)\;
         \(M_{\text{next}}\leftarrow M_{\text{next}}\cup M_{\Delta}\)\;
         \(P_{\text{next}}(\varphi)\leftarrow P_{\text{next}}(\varphi)\cup P_{\Delta}(\varphi)\)\;
      }

      \((M_{\text{old}}, M_{\text{new}})\leftarrow (M_{\text{old}}\cup M_{\text{new}}, M_{\text{next}})\)\;
      \((P_{\text{old}}, P_{\text{new}})\leftarrow (P_{\text{old}}\cup P_{\text{new}}, P_{\text{next}})\)\;

@@

   \KwData{\begin{minipage}[t]{.8\textwidth}
     \strut finite set \(\Phi\) of polymorphic formulae \\
     finite set $S$ of substitutions\strut
     \end{minipage}}
   \KwResult{finite set of monomorphic formulae}

@@

   \ForEach{\(\varphi\in\Phi\) such that $\varphi$ is non-monomorphic}{
      \ForEach{\(\sigma\in \MonoSubst{\AllSubst, \(\TyVars{\(\varphi\)}\), \(\emptyset\), \(\{\}\)}\)}{
         \textcolor{ourblueviolet}{
         \eIf{\(|\NewFormulae|<\Limit\)}{
            \textcolor{black}{
            \(\NewFormulae\leftarrow\NewFormulae\cup \{\sigma(\varphi)\}\)\;}
         }
         {\Return \NewFormulae}
      }}
   }

@@

   \Fn(){\FMonoStep{\(M_{\text{old}}, M_{\text{new}}, P_{\text{old}}(\varphi), P_{\text{new}}(\varphi)\)}}{
   \SetKwData{TypeVars}{type\_variables}
   \SetKwData{MaxMono}{max\_mono\_args}
   \SetKwData{MaxPoly}{max\_poly\_args}
   \SetKwData{UsedSubst}{\(S\)}

@@

   \KwData{\begin{minipage}[t]{.8\textwidth}
     \strut old and new monomorphic type argument maps \(M_{\text{old}}, M_{\text{new}}\) \\
     old and new non-monomorphic type argument maps \(P_{\text{old}}(\varphi), P_{\text{new}}(\varphi)\)\strut
      \end{minipage}
   }
   \KwResult{\begin{minipage}[t]{.8\textwidth}
      \strut monomorphic type argument map \\
      non-monomorphic type argument map \\
      set of substitutions\strut
      \end{minipage}
   }
   \BlankLine

@@

   \textcolor{ourblueviolet}{
   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M_{\text{old}}\cup M_{\text{new}}| \cdot \MonoMult\)}, \MonoCap}\;
   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P_{\text{old}}(\varphi)\cup P_{\text{new}}(\varphi)| \cdot \PolyMult\)}, \PolyCap}\;
   }

@@

   \ForEach{\(\sigma\in {}\)\SubstGen{\(M_{\text{new}}\), \(P_{\text{new}}(\varphi)\)}
\(\cup\) \SubstGen{\(M_{\text{new}}\), \(P_{\text{old}}(\varphi)\)}
\(\cup\) \SubstGen{\(M_{\text{old}}\), \(P_{\text{new}}(\varphi)\)}}{
      \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P_{\text{old}}(\varphi)\)}{
            \eIf{\TyVars{\((\upsilon_1, \dots, \upsilon_n)\)} \(\subseteq\) \Domain{\(\sigma\)}}{
               \textcolor{ourblueviolet}{
               \uIf{\(|M_{\text{next}}| < \MaxMono\) \label{algo:relative_bound}}{
                  \textcolor{black}{
                  \(M_{\text{next}}(f)\leftarrow M_{\text{next}}(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
                  \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;}
               }
               \uElseIf{\(|P_{\text{next}}(\varphi)| \geq \MaxPoly\) }{
                  \(M_{\text{next}} \leftarrow M_{\text{next}}\setminus (M_{\text{old}} \cup M_{\text{new}})\)\;
                  \(P_{\text{next}}(\varphi)\leftarrow P_{\text{next}}(\varphi)\setminus (P_{\text{old}}(\varphi)\cup P_{\text{new}}(\varphi))\)\;
                  \Return \((M_{\text{next}}, P_{\text{next}}(\varphi), \UsedSubst)\)
               }}
            }
            {
               \textcolor{ourblueviolet}{
               \If{\(|P_{\text{next}}(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
               \textcolor{black}{
               \(P_{\text{next}}(\varphi)(f)\leftarrow P_{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
               \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
            }}
            }

@@

   \Return \((M_{\text{next}}\setminus (M_{\text{old}} \cup M_{\text{new}}), P_{\text{next}}(\varphi)\setminus (P_{\text{old}}(\varphi)\cup P_{\text{new}}(\varphi)), \UsedSubst)\)
}

@@

   \(S\leftarrow \emptyset\)\;
   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
      \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
         \(\sigma\leftarrow \{\}\)\;
         \(i \leftarrow 1\)\;
         \While{\(i \leq n\)}{
            \eIf{\(\Match{\(\upsilon_i\), \(\tau_i\)}\) has the form \Success{\(\sigma_i\)} \And \Compatible{\(\sigma\), \(\sigma_i\)}}{
               \(\sigma\leftarrow\sigma_i\circ\sigma\)
            }
            {\Break}

@@

   }
   \Return \(S\)\;

}
\end{algorithm}
\end{quote}
\caption{Pseudocode for substitution generation}
\label{subst_gen}
\end{figure}

@@

   \SetKwFunction{MonoSubst}{mono\_substs}
   \SetKwFunction{Domain}{domain}
   \SetKwData{Limit}{subst\_limit}
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
   \SetKw{Let}{let}
   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

   \Fn(){\MonoSubst{\(S_{\text{all}}, S^\text{acc}, \sigma^\text{acc}, V\)}}{

   \KwData{%
      \hspace{-2.7em}\begin{minipage}[t]{.8\textwidth}
      \begin{enumerate}
         \item[] set of substitutions \(S_{\text{all}}\) and \(S^\text{acc}\)
         \item[] single substitution \(\sigma^\text{acc}\)
         \item[] set of type variables \(V\)
      \end{enumerate}
      \end{minipage}
   }

@@

   \eIf{\(V = \emptyset\)}{
    \Return \(S^\text{acc}\cup\{\sigma^\text{acc}\}\)\;%}
   }
   {
      \Let \(\alpha \) such that \(\alpha\in V\)\;
      \ForEach{\(\sigma_{\text{next}}\in S_{\text{all}}\) such that, \(\alpha\in\Domain{\(\sigma_{\text{next}}\)}\) \And \(\sigma^\text{acc}\) compatible with \(\sigma_{\text{next}}\)}{
         \textcolor{ourblueviolet}{
         \eIf{\(|S^\text{acc}|<\Limit\)}{
            \textcolor{black}{
            \(S^\text{acc}\leftarrow \MonoSubst{\(S_{\text{all}}, S^\text{acc}, \sigma_{\text{next}}\circ\sigma^\text{acc}, V\setminus\Domain{\(\sigma_{\text{next}}\)}\)}\)\;}
         }
         {\Return \(S^\text{acc}\)}}
      }
      \Return \(S^\text{acc}\)\;
   }
   }

@@

To monomorphise a polymorphic formula, we first compute its \emph{monomorphising substitutions} (\ref{mono_substs}). A substitution \(\sigma\) is a monomorphising substitution for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic. \ref{mono_substs} is a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S^\text{all}\), it will select a substitution \(\sigma_{\text{next}}\) from \(S^\text{all}\) which instantiates at least one of the type variables in \(V\). It is important that \(\sigma_{\text{next}}\) be compatible with \(\sigma^\text{acc}\) so that they may be composed and the function recursively called to instantiate the remaining type variables.

The Zipperposition implementation of this function makes use of a map from type variables to substitutions in place of a simple set to filter the relevant substitutions from \(S^\text{all}\) with greater ease. The \textcolor{ourblueviolet}{subst\_limit} exists for two main reasons:
\begin{enumerate}
   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{subst\_limit} new monomorphic clause per initial polymorphic clause. Generating an excessive number of new clauses can lead to the downstream prover being flooded.
   \item This function is the most explosive part of the algorithm. If \(S^\text{all}\) contains \(n\) substitutions which each instantiate exactly one of \(v\) different type variable, up to \(n^v\) monomorphising substititions may be generated. This should be considered in light of the fact that the total number of type variables (used to generate these substitutions) is already doubly exponential in the number of loop iterations.
\end{enumerate}

%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S^\text{acc}\) would represent the leaves that have already been visited and \(\sigma^\text{acc}\) would represent the current node of the tree.]

>>>>

\def\negvthinspace{\kern-0.083333em}
\def\vthinspace{\kern+0.083333em}

@@

\textbf{TODO: Mention that all sets referenced in the algorithm are finite.}

@@

   \KwData{set \(\Phi\) of polymorphic formulae}
   \KwResult{set of monomorphic formulae}

@@

   \((M_{\text{old}}, N_{\text{old}}) \leftarrow (\emptyset, \emptyset)\)\;
   \((M_{\text{new}}, N_{\text{new}}) \leftarrow \Init(\Problem)\)\;

@@

   \For{\(i = 1\) \To \textcolor{ourblueviolet}{\Loop}}{
      \((M_{\text{next}}, N_{\text{next}})\leftarrow(\emptyset,\emptyset)\)\;
      \ForEach{\(\varphi \in \Problem\)} {
         \((M_{\Delta}, N_{\Delta}(\varphi), S_{\Delta})\leftarrow \FStep(M_{\text{old}}, N_{\text{old}}(\varphi), M_{\text{new}}, N_{\text{new}}(\varphi))\)\;
         \(\AllSubst\leftarrow\AllSubst\cup S_{\Delta}\)\;
         \(M_{\text{next}}\leftarrow M_{\text{next}}\cup M_{\Delta}\)\;
         \(N_{\text{next}}(\varphi)\leftarrow N_{\text{next}}(\varphi)\cup N_{\Delta}(\varphi)\)\;
      }

      \((M_{\text{old}}, M_{\text{new}})\leftarrow (M_{\text{old}}\cup M_{\text{new}}, M_{\text{next}})\)\;
      \((N_{\text{old}}, N_{\text{new}})\leftarrow (N_{\text{old}}\cup N_{\text{new}}, N_{\text{next}})\)\;

@@

   \KwData{\begin{minipage}[t]{.8\textwidth}
     \strut set \(\Phi\) of polymorphic formulae \\
     set $S$ of substitutions\strut
     \end{minipage}}
   \KwResult{set of monomorphic formulae}

@@

   \ForEach{\(\varphi\in\Phi\) \textbf{such that} $\varphi$ is non-monomorphic}{
      \ForEach{\(\sigma\in \MonoSubst{\AllSubst, \(\TyVars{\(\varphi\)}\), \(\emptyset\), \(\{\}\)}\)}{
         \textcolor{ourblueviolet}{
         \eIf{\(|\NewFormulae|<\Limit\)}{
            \textcolor{black}{
            \(\NewFormulae\leftarrow\NewFormulae\cup \{\sigma(\varphi)\}\)\;}
         }
         {\Return \NewFormulae}
      }}
   }

@@

   \Fn(){\FMonoStep{\(\varphi, M_{\text{old}}, M_{\text{new}}, N_{\text{old}}(\varphi), N_{\text{new}}(\varphi)\)}}{
   \SetKwData{TypeVars}{type\_variables}
   \SetKwData{MaxMono}{max\_mono\_args}
   \SetKwData{MaxPoly}{max\_poly\_args}
   \SetKwData{UsedSubst}{\(S\)}

@@

   \KwData{\begin{minipage}[t]{.8\textwidth}
     \strut polymorphic formula $\varphi$ \\
     old and new monomorphic type argument maps \(M_{\text{old}}, M_{\text{new}}\) \\
     old and new non-monomorphic type argument maps \(N_{\text{old}}(\varphi), N_{\text{new}}(\varphi)\)\strut
      \end{minipage}
   }
   \KwResult{\begin{minipage}[t]{.8\textwidth}
      \strut monomorphic type argument map \\
      non-monomorphic type argument map \\
      set of substitutions\strut
      \end{minipage}
   }
   \BlankLine

@@

   \textcolor{ourblueviolet}{
   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M_{\text{old}}\cup M_{\text{new}}| \cdot \MonoMult\)}, \MonoCap}\;
   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)| \cdot \PolyMult\)}, \PolyCap}\;
   }

@@

   \ForEach{\(\sigma\in {}\)\SubstGen{\(M_{\text{new}}\), \(N_{\text{new}}(\varphi)\)}
\(\cup\) \SubstGen{\(M_{\text{new}}\), \(N_{\text{old}}(\varphi)\)}
\(\cup\) \SubstGen{\(M_{\text{old}}\), \(N_{\text{new}}(\varphi)\)}}{
      \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in N_{\text{old}}(\varphi)\)}{
            \eIf{\TyVars{\((\upsilon_1, \dots, \upsilon_n)\)} \(\subseteq\) \Domain{\(\sigma\)}}{
               \textcolor{ourblueviolet}{
               \uIf{\(|M_{\text{next}}| < \MaxMono\) \label{algo:relative_bound}}{
                  \textcolor{black}{
                  \(M_{\text{next}}(f)\leftarrow M_{\text{next}}(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
                  \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;}
               }
               \uElseIf{\(|N_{\text{next}}(\varphi)| \geq \MaxPoly\) }{
                  \(M_{\text{next}} \leftarrow M_{\text{next}}\setminus (M_{\text{old}} \cup M_{\text{new}})\)\;
                  \(N_{\text{next}}(\varphi)\leftarrow N_{\text{next}}(\varphi)\setminus (N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi))\)\;
                  \Return \((M_{\text{next}}, N_{\text{next}}(\varphi), \UsedSubst)\)
               }}
            }
            {
               \textcolor{ourblueviolet}{
               \If{\(|N_{\text{next}}(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
               \textcolor{black}{
               \(N_{\text{next}}(\varphi)(f)\leftarrow N_{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
               \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
            }}
            }

@@

   \Return \((M_{\text{next}}\setminus (M_{\text{old}} \cup M_{\text{new}}), N_{\text{next}}(\varphi)\setminus (N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)), \UsedSubst)\)
}

@@

   \(S\leftarrow \emptyset\)\;

   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
      \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
         \(\sigma\leftarrow \{\}\)\;
         \(i \leftarrow 1\)\;
         \While{\(i \leq n\)}{
            \eIf{\(\Match{\(\upsilon_i\), \(\tau_i\)}\) has the form \Success{\(\sigma_i\)} \And \Compatible{\(\sigma\), \(\sigma_i\)}}{
               \(\strut\sigma\leftarrow\sigma_i\circ\sigma\)
            }
            {\Break}

@@

   \Return \(S\)\;

}
\end{algorithm}
\end{quote}
\caption{Pseudocode for match generation}
\label{subst_gen}
\end{figure}

@@

   \SetKwFunction{Compatible}{compatible}
   \SetKwFunction{MonoSubst}{mono\_substs}
   \SetKwFunction{Domain}{domain}
   \SetKwData{Limit}{max\_substs}
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
   \SetKw{Let}{let}
   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

   \Fn(){\MonoSubst{\(S, V, S_{\text{res}}, \sigma\)}}{

   \KwData{\begin{minipage}[t]{.8\textwidth}
     \strut set \(S\) of substitutions \\
     set \(V\) of type variables \\
     \strut set \(S_{\text{res}}\) of substitutions \\
     substitution \(\sigma\)\strut
      \end{minipage}
   }

@@

   \eIf{\(V = \emptyset\)}{
    \Return \(S_{\text{res}}\cup\{\sigma\}\)\;%}
   }
   {
      \Let \(\alpha \) \textbf{such that} \(\alpha\in V\)\;
      \ForEach{\(\sigma_{\negvthinspace\Delta}\in S\) \textbf{such that} \(\alpha\in\Domain{\(\sigma_{\negvthinspace\Delta}\)}\) \And \(\Compatible(\sigma,\sigma_{\negvthinspace\Delta})\)}{
         \textcolor{ourblueviolet}{
         \eIf{\(|S_{\text{res}}|<\Limit\)}{
            \textcolor{black}{
            \(S_{\text{res}}\leftarrow \MonoSubst{\(S, V\setminus\Domain{\(\sigma_{\negvthinspace\Delta}\)}, S_{\text{res}}, \sigma_{\negvthinspace\Delta}\circ\sigma\)}\)\;}
         }
         {\Return \(S_{\text{res}}\)}}
      }
      \Return \(S_{\text{res}}\)\;
   }
   }

@@

To monomorphise a polymorphic formula, we first compute its \emph{monomorphising substitutions} (\ref{mono_substs}). A substitution \(\sigma\) is a monomorphising substitution for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic. \ref{mono_substs} is a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S^\text{all}\), it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S^\text{all}\) which instantiates at least one of the type variables in \(V\). It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they may be composed and the function recursively called to instantiate the remaining type variables.

The Zipperposition implementation of this function makes use of a map from type variables to substitutions in place of a simple set to filter the relevant substitutions from \(S^\text{all}\) with greater ease. The \textcolor{ourblueviolet}{max\_substs} exists for two main reasons:
\begin{enumerate}
   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic clause per initial polymorphic clause. Generating an excessive number of new clauses can lead to the downstream prover being flooded.
   \item This function is the most explosive part of the algorithm. If \(S^\text{all}\) contains \(n\) substitutions which each instantiate exactly one of \(v\) different type variable, up to \(n^v\) monomorphising substititions may be generated. This should be considered in light of the fact that the total number of type variables (used to generate these substitutions) is already doubly exponential in the number of loop iterations.
\end{enumerate}

%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]
